### 2018-06-08 latex/egsr17-cgf/section_results.tex
revision 2

Our computer algebra framework and the precomputation has been implemented in Python. The runtime component of our solver has been implemented in C++. We use a naive implementation of a CG solver, which has been modified such that we do not need to expliticly compute the normal form of the linear system to solve. We use the sparse matrix representation and sparse matrix vector product from the Eigen linear algebra library.

The solver for classical diffusion is based on the diffusion equation, which is derived by collapsing the $P_1$-equations:
\begin{align} 
\nabla\left(\frac{1}{3\sigma_t}\nabla L^{0,0}\right)  = -Q^{0,0}
\label{eq:cda}
\end{align}

Since our solver can work with any PDE, which results in a linear system of equations, we put Equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which would generate the correct stencil code automatically.

@@

First we ran our solver on the 2D checkerboard, a very common test case in other fields. The problem has dimensions $7\times 7$ and is discretized with resolution $71\times 71$. Unit size blocks are filled with purely absorbing medium with $\sigma_a=10$ in a checkerboard pattern. All other blocks are filled with a purely scattering medium with $\sigma_s=1$.

Solving the standard checkerboard problem, allows us to validate our solver against the solver from Seibold et al.~\cite{Seibold14}, which solves for the time-dependent and complex-valued $P_N$-equations in the 2D case only. The 2D case is derived by assuming that all SH coefficients, RTE parameters and boundary conditions are z-independent. This causes all SH coefficients and moment equations for which $l+m$ is odd to vanish. Due to the time-dependency, their approach is to do explicit incremental steps in time. We run their solver for many timesteps, in order to get a result which is close to steady state.

\begin{figure}[h]
\centering
\begin{subfigure}{0.49\columnwidth}
%\missingfigure{test}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered_starmap.png}
\end{subfigure}%
\hspace{0.01\columnwidth}
\begin{subfigure}{0.49\columnwidth}
%\missingfigure{test2}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered.png}
\end{subfigure}%
\vspace{-0.2in}
\icaption{Comparison of the result for the checkerboard test using Starmaps time-stepping solver~\cite{Seibold14} (left) against our steady-state solver (right) with $N=5$. Our results are in good agreement.}
\label{fig:vs_starmap}
\end{figure}

@@

\subsection{Pointsource problem}

@@

For our test case, we choose a FD resolution of $80\times80\times80$, an extinction coefficient $\sigma_t=8.0$ and albedo $\alpha=0.9$. We run the solver for different truncation values $N$. In Figure~\ref{fig:pointsource}~\subref{fig:pointsource_pn}, we see that the solution becomes increasingly accurate for higher truncation order. The ground truth is underestimated when $N$ is odd, and overestimated when $N$ is even. We further see, that the $P_1$ solution exactly matches the results from CDA, which confirmes, that the latter is only a collapsed version of the former. The time to solve is significant with $10m$ for $P_1$ and $45m$ with $P_5$. With these performance characteristics, the $P_N$-method is clearly not competitive in comparison with much diffusion and FLD solver, which are much faster.

@@

Finally, we run our solver on a procedural cloud dataset to get an idea of its performance in more practical applications. Figure~\ref{fig:teaser}~\subref{fig:nebulae_ours} shows the result of $P_5$ for a procedurally generated heterogeneous cloud with an isotropic phase function. We see that at order $N=5$, our method can capture indirect illumination similarily well as FLD and is significantly better than CDA as expected. The indirectly illuminated region at the bottom appears to be closer to the path-traced result as opposed to the solution form FLD which is very flat in that region. However, in many other areas, $P_5$ seems to still suffer a lot from the problem of energy loss near transitions from dense regions to vacuum. It appears that going higher order a few steps, mitigates this problem only slowly at an immense cost of compute and storage.

The main characteristic of the nebulae dataset is the presence of vacuum. We found that having vacuum regions in the dataset will cause the condition number to become infinite and the solver practically doesnt converge. We therefore introduced a minimum threshold for the extinction coefficient $\sigma_t$. Every voxel with an extinction coefficient smaller than the threshold would be set to the threshold value. In Figure~\ref{fig:results_convergence} we show the effect of the minimum threshold on the convergence behavior. As the threshold increases, convergence improves.

@@

\begin{figure}[h]
\centering
%\missingfigure{test}
\includegraphics[width=0.9\columnwidth]{figures/fig_nebulae_p1_convergence.pdf}
%\vspace{-0.2in}
\vspace{-0.1in}
\icaption{Convergence behavior of our solver with $N=1$ for the nebulae dataset and for varying minimum thresholds of the extinction coefficient $\sigma_t$. Threshold values and an estimate for the condition number of $A$ (Matlabs condest function) are shown next to the plots. The convergence deteriorates as the threshold decreases. Once it reaches zero, the presence of pure vaccuum makes the condition number infinite.}
\label{fig:results_convergence}
\end{figure}

>>>>

Our computer algebra framework and the precomputation has been implemented in Python. The runtime component of our solver has been implemented in C++. We use a naive implementation of a CG solver, which has been modified such that we do not need to expliticly compute the normal form of the linear system to solve. We use the sparse matrix representation and sparse matrix vector product from the Eigen linear algebra library (eigen.tuxfamily.org).

The solver for classical diffusion is based on the diffusion equation, which is derived by collapsing the $P_1$-equations:
\begin{align} 
\nabla\left(\frac{1}{3\sigma_t}\nabla L^{0,0}\right)  = -Q^{0,0}
\ .
\label{eq:cda}
\end{align}

Since our solver can work with any PDE, which results in a linear system of equations, we put Equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which generates the correct stencil code automatically.

@@

First we ran our solver on the 2D checkerboard, a very common test case in other fields. The problem has dimensions $7\times 7$ and is discretized with resolution $71\times 71$. Unit size blocks are filled with purely absorbing medium $\sigma_a=10$ in a checkerboard pattern. All other blocks are filled with a purely scattering medium with $\sigma_s=1$.

Solving the standard checkerboard problem, allows us to validate our solver against the solver from Seibold et al.~\cite{Seibold14}, which solves for the time-dependent and complex-valued $P_N$-equations in the 2D case only. The 2D case is derived by assuming that all SH coefficients, RTE parameters and boundary conditions are z-independent. This causes all SH coefficients and moment equations for which $l+m$ is odd to vanish. Due to the time-dependency, their approach is to do explicit incremental steps in time. We run their solver for many timesteps to get a result close to steady state.

\begin{figure}[h]
\centering
\begin{subfigure}{0.49\columnwidth}
%\missingfigure{test}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered_starmap.png}
\end{subfigure}%
\hspace{0.01\columnwidth}
\begin{subfigure}{0.49\columnwidth}
%\missingfigure{test2}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered.png}
\end{subfigure}%
\vspace{-0.2in}
\icaption{Comparison of the result for the checkerboard test using StaRMAP's time-stepping solver~\cite{Seibold14} (left) against our steady-state solver (right) with $N=5$. Our results are in good agreement.}
\label{fig:vs_starmap}
\end{figure}

@@

\subsection{Point source problem}

@@

For our test case, we choose a FD resolution of $80\times80\times80$, an extinction coefficient $\sigma_t=8.0$ and albedo $\alpha=0.9$. We run the solver for different truncation values $N$. In Figure~\ref{fig:pointsource}~\subref{fig:pointsource_pn}, we see that the solution becomes increasingly accurate for higher truncation order. The ground truth is underestimated when $N$ is odd, and overestimated when $N$ is even. We further see that the $P_1$ solution exactly matches the results from CDA, which confirms that the latter is only a collapsed version of the former. The time to solve is significant with $10m$ for $P_1$ and $45m$ with $P_5$. With these performance characteristics, the $P_N$-method is clearly not competitive in comparison with CDA and FLD solver, which are much faster.

@@

Finally, we run our solver on a procedural cloud dataset to get an idea of its performance in more practical applications. Figure~\ref{fig:teaser}~\subref{fig:nebulae_ours} shows the result of $P_5$ for a procedurally generated heterogeneous cloud with an isotropic phase function. We see that at order $N=5$, our method can capture indirect illumination similarily well as FLD and is significantly better than CDA as expected. The indirectly illuminated region at the bottom appears to be closer to the path-traced result as opposed to the solution from FLD which is very flat in that region. However, in many other areas, $P_5$ seems to still suffer a lot from the problem of energy loss near transitions from dense regions to vacuum. It appears that going higher order a few steps mitigates this only slowly at a high cost of compute and storage.

The main characteristic of the nebulae dataset is the presence of vacuum. We found that having vacuum regions in the dataset will cause the condition number to become infinite and the solver practically does not converge. We therefore introduced a minimum threshold for the extinction coefficient $\sigma_t$. Every voxel with an extinction coefficient smaller than the threshold is set to the threshold value. In Figure~\ref{fig:results_convergence} we show the effect of the minimum threshold on the convergence behavior. As it increases, convergence improves.

@@

\begin{figure}[h]
\centering
%\missingfigure{test}
\includegraphics[width=0.9\columnwidth]{figures/fig_nebulae_p1_convergence.pdf}
%\vspace{-0.2in}
\vspace{-0.1in}
\icaption{Convergence behavior of our solver with $N=1$ for the nebulae dataset and for varying minimum thresholds of the extinction coefficient $\sigma_t$. Threshold values and an estimate for the condition number of $A$ (Matlabs condest function) are shown next to the plots. The convergence deteriorates as the threshold decreases. Once it reaches zero, the presence of pure vacuum makes the condition number infinite.}
\label{fig:results_convergence}
\end{figure}

### 2018-06-08 latex/egsr17-cgf/solver.tex
revision 2

The truncation order $N$ is the key input parameter to the solver. With higher values, manual implementation of the solver from the equations would be ardous, error-prone and time-consuming. We therefore decided to make use of the computer algebra representation and designed our solver around it.

@@

\begin{figure}[b]
\centering
%\missingfigure{test}
\includegraphics[width=0.7\columnwidth]{figures/fig_stencil_pipeline.pdf}
%\vspace{-0.2in}
\icaption{Creating the stencil code requires several steps, usually done by hand. We express the given problem in a computer algebra representation and use it to fully automate the process.}
\label{fig:stencile_pipeline}
\end{figure}

@@

Stencil generation entails discretizing a PDE at a hypothetical center voxel $(i, j, k)$ (assuming to mean the voxel center most of the time). Finite differences create weighted references to other voxels (e.g. $i+1, j, k$). After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and offsets (Figure~\ref{fig:stencile_pipeline}). $(i, j, k)$ will only be known during runtime, when the stencil is executed for a particular unknown (row). Then the offsets can be used to find the component index into the solution-vector, and weights can be evaluated for concrete world space position. We refer to the space with the hypthetical voxel $(i, j, k)$ at the center as stencil space.

@@

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.2in}
\includegraphics[width=0.6\columnwidth]{figures/fig_discretization.pdf}
%\end{center}
\end{wrapfigure}The spatial discretization is done by parsing the expression tree from the root. The discrete substitute for the continuous position variable $\vec{x}$ is initialized with $ijk$. Differential operator nodes are replaced by a subtree, which expresses the finite difference approximation (including voxelsize factor $h$). The subtree of the differential operator node is duplicated for different offsets to the discrete variable $(i, j, k)$. Since this offset only applies to the substree, a stack is maintained by the parser to manage scope. Whenever the parser encounters the continuous variable $\vec{x}$ in the expression tree, its node in the tree is replaced by the discrete substitute, currently on top of the stack. Nested differential operators yield higher order finite difference stencils as expected.

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.2in}
\includegraphics[width=0.6\columnwidth]{figures/fig_codegen.pdf}
%\end{center}
\end{wrapfigure}
Factorization into canonical form is done as a manipulation pass on the mathematical expression tree. The result allows to implement the code generation in a straight forward fashion. For each term, the $ijk$-offset is retrieved from the unknown. The factor-expression, which is multipled with the unknown, is extracted from the tree and used to generate source code for evaluating the factor-expression during runtime (including calls for evaluating RTE-parameter fields).

@@

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.1in}
\includegraphics[width=0.6\columnwidth]{figures/fig_staggered_interpolation.pdf}
\vspace{-.1in}
%\end{center}
\end{wrapfigure}Our solver supports placement of coefficients at arbitrary staggered grid locations (see Figure~\ref{fig:staggeredgrid}). This means, that during the discretization step, the discrete location $(i, j, k)$ (at which the unknown is meant to be evaluated) might not coincide with the location of the unknown. To solve this, depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at $(i, j, k)$ from their defined locations. If those happen to coincide, the coefficient itsself is returned. This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center.

@@

The stencil code is generated once for every value of $N$ and compiled with the runtime component of our solver. The runtime executes the stencil for every voxel to propagate the system matrix $A$ and RHS vector $\vec{Q}$ with numerical values.

The number of rows is determined by the number of voxels times the number of coefficients per voxel (see Figure~\ref{fig:matrix_layout}) and can therefore become very large for high resolution and high truncation order. The matrix $A$ is square and extremely sparse, due to the local structure of the finite differences discretization. Unfortunately, it is non-symmetric due to the transport term and not diagonal dominant, which rules out many \rev{iterative} methods for solving linear systems. \rev{Iterative methods are useful, as they allow balancing accuracy against performance by tweaking the convergence threshold.} We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead. This gives a symmetric and positive definit system matrix $A^TA$, albeit with a higher condition number. Investigation of other solution schemes (e.g. multigrid) would be an interesting avenue for future work. \rev{However, more importantly, in the presence of vacuum regions, the matrix $A$ becomes singular and the system cannot be solved at all. This requires the introduction of a minimum threshold for the extinction coefficient $\sigma_t$.}

Our solver supports Neumann and Dirichlet boundary conditions. They are handled transparently by the code which generates the stencil. Whenever the stencil accumulates coefficients into a boundary location, the framework either ignores the write operation (Dirichlet BC) or accumulates into the row and column in $A$ of the coefficient in the closest voxel inside the domain (Neumann BC). This is done changing the index of the written component under the hood.

>>>>

The truncation order $N$ is the key input parameter to the solver. With higher values, manual implementation of the solver from the equations would be arduous, error-prone and time-consuming. We therefore decided to make use of the computer algebra representation and designed our solver around it.

@@

@@

Stencil generation entails discretizing a PDE at a hypothetical center voxel $(i, j, k)$ (assumed to mean the voxel center most of the time). Finite differences create weighted references to other voxels (e.g. $i+1, j, k$). After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and offsets (Figure~\ref{fig:stencile_pipeline}). Voxel $(i, j, k)$ will only be known during runtime, when the stencil is executed for a particular unknown (row). Then the offsets can be used to find the component index into the solution-vector, and weights can be evaluated for concrete world space position. We refer to the space with the hypothetical voxel $(i, j, k)$ at the center as stencil space.

\vspace{0.1in}

\begin{figure}[h]
\centering
%\missingfigure{test}
\includegraphics[width=0.7\columnwidth]{figures/fig_stencil_pipeline.pdf}
%\vspace{-0.2in}
\icaption{Creating the stencil code requires several steps, usually done by hand. We express the given problem in a computer algebra representation and use it to fully automate the process.}
\label{fig:stencile_pipeline}
\end{figure}

@@

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.2in}
\includegraphics[width=0.6\columnwidth]{figures/fig_discretization.pdf}
%\end{center}
\end{wrapfigure}The spatial discretization is done by parsing the expression tree from the root. The discrete substitute for the continuous position variable $\vec{x}$ is initialized with $ijk$. Differential operator nodes are replaced by a subtree, which expresses the finite difference approximation (including voxelsize factor $h$). The subtree of the differential operator node is duplicated for different offsets to the discrete variable $(i, j, k)$. Since this offset only applies to the subtree, a stack is maintained by the parser to manage scope. Whenever the parser encounters the continuous variable $\vec{x}$ in the expression tree, its node in the tree is replaced by the discrete substitute, currently on top of the stack. Nested differential operators yield higher order finite difference stencils as expected.

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.2in}
\includegraphics[width=0.6\columnwidth]{figures/fig_codegen.pdf}
%\end{center}
\end{wrapfigure}
Factorization into canonical form is done as a manipulation pass on the mathematical expression tree. The result allows us to implement the code generation in a straightforward fashion. For each term, the $ijk$-offset is retrieved from the unknown. The factor-expression, which is multipled with the unknown, is extracted from the tree and used to generate source code for evaluating the factor-expression during runtime (including calls for evaluating RTE-parameter fields).

@@

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.1in}
\includegraphics[width=0.6\columnwidth]{figures/fig_staggered_interpolation.pdf}
\vspace{-.1in}
%\end{center}
\end{wrapfigure}Our solver supports placement of coefficients at arbitrary staggered grid locations (see Figure~\ref{fig:staggeredgrid}). This means, that during the discretization step, the discrete location $(i, j, k)$ (at which the unknown is meant to be evaluated) might not coincide with the location of the unknown. To solve this, depending on how the two are located relativeto each other, the parser returns an expression which interpolates the coefficients at $(i, j, k)$ from their defined locations. If those happen to coincide, the coefficient itsself is returned. This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center.

@@

The stencil code is generated once for every value of $N$ and compiled with the runtime component of our solver. The runtime executes the stencil for every voxel to populate the system matrix $A$ and RHS vector $\vec{Q}$ with numerical values.

The number of rows is determined by the number of voxels times the number of coefficients per voxel (see Figure~\ref{fig:matrix_layout}) and can therefore become very large for high resolution and high truncation order. The matrix $A$ is square and extremely sparse, due to the local structure of the finite differences discretization. Unfortunately, it is non-symmetric due to the transport term and not diagonal dominant, which rules out many \rev{iterative} methods for solving linear systems. \rev{Iterative methods are useful, as they allow balancing accuracy against performance by tweaking the convergence threshold.} We address this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead. This gives a symmetric and positive definite system matrix $A^TA$, albeit with a higher condition number. Investigation of other solution schemes (e.g. multigrid) would be an interesting avenue for future work. \rev{However, more importantly, in the presence of vacuum regions, the matrix $A$ becomes singular and the system cannot be solved at all. This requires the introduction of a minimum threshold for the extinction coefficient $\sigma_t$.}

Our solver supports Neumann and Dirichlet boundary conditions. They are handled transparently by the code which generates the stencil. Whenever the stencil accumulates coefficients into a boundary location, the framework either ignores the write operation (Dirichlet BC) or accumulates into the row and column in $A$ of the coefficient in the closest voxel inside the domain (Neumann BC). This is done by changing the index of the written component.
